---
title: "Basic Data Wrangling"
author: "sbsambado"
date: "8/11/2021"
output: 
  html_document:
    toc: true
    toc_depth: 1
    toc_float: true
---

This file is intended to be an intro to data wrangling and tidyverse for summer programming at UCSB in summer 2021.


Workshop Agenda (8/12/21)

1. The Big Picture: Let's think about data
2. Tidyverse
3. Examples

# 1. The Big Picture
## 1. Data visualization is meant to convey a story. 
There are many ways to tell a story, but there are some general guidelines on how to build that story

  + What is your outcome of interest - your **dependent** variable 'y'? 
      + That outcome should be placed on the y-axis
  
  + What is influencing your outcome of interest - your **independent** variable 'x'? 
    + That should be placed on the x-axis
  
## 2. Your type of data determines how you shape your data 
        +  Numeric: numbers
        
        +  Categorical: words
        
        +  Continuous: numbers can take on **any** value
        
        +  Discrete: numbers or words can only be **certain** values

## 3. Plotting basics
    + If you're only plotting 1 numeric variable --> histogram (ggplot geom: geom_histogram)
    
    + If y and x are **both** numeric --> scatterplot (ggplot geom: geom_point)
    
    + If y is numeric, and x is categorical --> boxplot (ggplot geom: geom_boxplot)

# 3. Data visualization is the **first** AND **last** step of statistical analysis

  +  Be prepared to make a lot of changes along the way & have fun with it!
  
  + Figures will never be 100% perfect in the creator's eyes, but if it tells your story in a **clear and effective** way, then that is a job *well done*.

Okay, end philosophical ramble. Let's get started.    


Here are helpful links I derive a lot of my information about tidyverse & `dplyr` : 

[Basic Tidyverse Concepts](https://homerhanumat.github.io/r-notes/tidyverse-concepts.html)


[Tidyverse with animations](https://www.garrickadenbuie.com/project/tidyexplain/)


[Tidyverse-cookbook](https://rstudio-education.github.io/tidyverse-cookbook/tidy.html)


[YaRrr! The pirate’s guide to R -- chapter 10.4](https://bookdown.org/ndphillips/YaRrr/dplyr.html)
*I also think this is an excellent guide for beginners to R!*



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE) # this means

## necessary libraries
# remember, install.packages("PACAKGE") then library(PACKAGE)
# you only need to install a package once to call the library
# after you install.packages(), then hashtag # install.packages()

# install.packages("tidyverse")
library(tidyverse)
# install.packages("devtools")
library("devtools")
# install.packages("mosaicData")
library(mosaicData)

## two packages to accompany the text
devtools::install_github("homerhanumat/bcscr")
devtools::install_github("homerhanumat/tigerData")


```

## *tidyverse isn't a package, it's a collection of packages* 
**just like universities are made up of colleges that have specific functions**

### Tidy data has the following rules

+ 1. Each **variable** has its own **column**

+ 2. Each **observation** has its own **row**

+ 3. Each **value** has its own **cell**

#### Basics

**1. pipe operator `%>` : chains function together**

```{r}
# using pipe operator
"hello" %>% rep(times = 4)

# without pipe operator
rep("hello", times = 4)
```

**2. tibble : type of dataframe suited for tidyverse**

+ good for large datasets
```{r}
# dataframe
survey <- bcscr::m111survey # uploading a 
class(survey)

# convert to tibble
survey <- as_tibble(survey)
class(survey)
```


**3. dplyr package: tool for data manipulation** 

+ a. `filter()` and `select()` for sub-setting

+ b. `mutate()` for transforming variables

+ c. `group_by()` and `summarise()` for numerical summaries


a. examples of **`filter()` and `select()`** to chose optimal observations & variables
```{r}
# usually filter() and select() are your first arguments when you get a dataset
survey %>% # call tibble
  filter((sex == "male" & height > 70) | (sex == "female" & height < 55)) %>% # filter picks out rows/observations ## we're selecting observations that are men > 70 in tall AND women < 55 in tall
  select(sex, height, fastest) # selects picks out columns/variables
# let's leave out columns/variables ; use `-`
survey %>%
  select(-ideal_ht, -love_first) # still have lots of columns left!
# want to see how many observations you have left post subsetting?
survey %>%
  filter(love_first == "yes" & fastest > 120) %>%
  nrow() # count the rows/observations left
```

b. examples of **`mutate()`** to transform variables
```{r}
# keep new variable name on LEFT side of `=`
# keep function on RIGHT side of `=` that depends on the current value in the tibble
survey %>%
  mutate(dareDevil = fastest > 125) %>% # make new column/variable called dareDevil ## this will provide TRUE (> 125) /FALSE (< 125) for each observation with regards to its fastest value
  select(sex, fastest, dareDevil) # select desired variables
# You can double dip mutations within one line of code
survey %>%
  mutate(dareDevil = fastest > 125, 
         speedKmHR = fastest * 1.60934) %>% # made two new variables/columns based on previous data in column 'fastest')
  ggplot(aes(x = dareDevil, y = GPA)) + # plot these new variable values
  geom_boxplot(fill = 'burlywood', outlier.alpha = 0) + # using boxplot to visualize datat
  geom_jitter(width = 0.2) # avoid points overlapping and making them a certain size
  
```



c. examples of **`group_by()` and `summarise()`** to generate numerical summaries 

+ *^^ my most used functions in dplyr*

```{r}
CPS85 %>% # tibble that comes from the `mosaicData`
  group_by(sex) %>% # divide observations based on their values/observations in the variable 
  summarize(meanWage = mean(wage)) # make new tibble that has mean wage for each level of sex, you will only have two columns/variables in this tibble 1) sex and 2) meanWage
# create more than 1 summary variable
CPS85 %>% 
  group_by(sex) %>% 
  summarize(meanWage = mean(wage),
            n = n()) # n() is the count function, very helpful for most summaries ## will count total observations for your variable meanWage
```

**other summaries you can do are:**

+ minimum value
+ first quartile (q1)
+ median 
+ third quartile (q3)
+ max

if you want to run all 5 summaries (which you should), use `fivenum()` function
```{r}
# summaries for entire tibble (men and women included)
CPS85 %>%
  .$wage %>% # select one column/vector of information you want to summarize
  fivenum() # run all 5 summary statistics
# summaries broken down for each sex (more appropriate of comparisons!)
CPS85 %>%
  group_by(sex) %>% 
  summarise(n = n(), # count observations
            min = fivenum(wage)[1],
            Q1 = fivenum(wage)[2],
            median = fivenum(wage)[3],
            Q3 = fivenum(wage)[4],
            max = fivenum(wage)[5])
# you can also compare multiple variables by group
CPS85 %>% 
  group_by(sector, sex) %>%  # looking at SEX and SECTOR
  summarise(n = n(),
            min = fivenum(wage)[1],
            Q1 = fivenum(wage)[2],
            median = fivenum(wage)[3],
            Q3 = fivenum(wage)[4],
            max = fivenum(wage)[5])
```


# More practice!

```{r}
data("flights", package = "nycflights13")
```

**filter out NA cases**

+ Want to correctly count number of flights that left airport
```{r}
flights %>% 
  filter(!is.na(dep_delay)) %>% 
  group_by(origin) %>%
  summarise(departures = n(),
            meanDelay = mean(dep_delay))
```

**filter for a specific date**

+ Find flights that occurred on June 26
```{r}
flights %>% 
  filter(month == 6 & day == 26) %>% 
  ggplot(aes(x = origin, y = distance)) +
    geom_violin(fill = "burlywood") +
    geom_jitter(width = 0.25, size = 0.1)
```

**filter for specific date, variable value and types of variables**

+ Find flights that occurred on June 26 that were over 4000 mi and chose only relevant information of origin, destination, and distance
```{r}
flights %>% 
  filter(month == 6 & day == 26 & distance > 4000) %>% 
  select(origin, dest, distance)

```

